Covariate analysis of CMC RNASeq data reprocessed with STAR-feature counts workflow with Ensembl v75 (DLPFC with CQN normlaisation)

### 2017-09-16 code/Rmd/CMC_DLPFC_STAR_ENSEMBL_v75_CQN.Rmd
Added markdown to process DLPFC samples in CMC

---
title: "Covariate analysis of CMC RNASeq data reprocessed with STAR-feature counts workflow with Ensembl v75 (ACC and DLPFC with CQN normlaisation)"
author: "Thanneer Perumal"
date: "`r date()`"
output: html_document
---
```{r knit2synapse, eval=FALSE}
library(synapseClient)
library(knit2synapse)

@@

knit2synapse::knitToFolderEntity(file = "./CMC_ACC_DLPFC_STAR_ENSEMBL_v75_CQN.Rmd",
                                 parentId = "syn9872607",
                                 entityName = 'ACC and DLPFC - STAR - FEATURE COUNTS - ENSEMBL v75 - CQN')
```

@@

```{r synapse.parameters, include=FALSE, cache=TRUE}
parentId = "syn9872607";
activityName = 'Covariate adjustments';
activityDescription = 'Covariate analysis of STAR aligned feature counts with Ensembl v75 with CQN normalisation (ACC and DLPFC)';

thisFileName <- 'CMC_ACC_DLPFC_STAR_ENSEMBL_v75_CQN.Rmd'

# Github link
thisRepo <- getRepo(repository = "th1vairam/Brain_Reg_Net", ref="branch", refName='CMC')
thisFile <- getPermlink(repository = thisRepo, repositoryPath=paste0('code/Rmd/',thisFileName))
```
### Data download
Obtain count matrix and metadata from synapse.
```{r download.data, cache=TRUE}
# Download reprocessed counts (ACC)
COUNT_ACC_ID = 'syn8414424'
ALL_USED_IDs = COUNT_ACC_ID
COUNT_ACC = downloadFile(COUNT_ACC_ID)

# Download reprocessed counts (DLPFC)
COUNT_DLPFC_ID = 'syn8413221';
ALL_USED_IDs[length(ALL_USED_IDs)+1] = COUNT_DLPFC_ID
COUNT_DLPFC = downloadFile(COUNT_DLPFC_ID)

# Join ACC and DLPFC together
COUNT = join_all(list(COUNT_ACC, COUNT_DLPFC), type = 'full') %>%
  as.data.frame()

GENE.LEN = dplyr::select(COUNT_ACC, Geneid, Length) %>%
  dplyr::rename(ensembl_gene_id = Geneid) %>%
  unique()
rownames(GENE.LEN) = GENE.LEN$ensembl_gene_id

@@

# Get RNA qc metrics from synapse
METADATA_QC_ACC_ID = 'syn2929053'
ALL_USED_IDs[length(ALL_USED_IDs)+1] = METADATA_QC_ACC_ID
METADATA_QC_ACC = read.table(synGet(METADATA_QC_ACC_ID)@filePath, fill = T, header = T, sep = ',') %>%
  lapply(as.character)

# Get RNA qc metrics from synapse
METADATA_QC_DLPFC_ID = 'syn2279446'
ALL_USED_IDs[length(ALL_USED_IDs)+1] = METADATA_QC_DLPFC_ID
METADATA_QC_DLPFC = read.table(synGet(METADATA_QC_DLPFC_ID)@filePath, fill = T, header = T, sep = ',') %>%
  lapply(as.character)

METADATA_QC = rbindlist(list(METADATA_QC_ACC, METADATA_QC_DLPFC), use.names = T, fill = T) %>%
  dplyr::filter(is.na(Exclude.)) %>%
  dplyr::select(Sample.RNA.ID, Ribozero.Batch, Library.Batch, Flowcell.Batch, Mapped.Reads, 
                Intragenic.Rate, Exonic.Rate, Intronic.Rate, Intergenic.Rate, Genes.Detected, 
                Transcripts.Detected, Expression.Profiling.Efficiency, rRNA.Rate, Total.Reads, Percent.Aligned) %>%
  dplyr::rename(SampleID = Sample.RNA.ID, RibozeroBatch = Ribozero.Batch, LibraryBatch = Library.Batch, 
                FlowcellBatch = Flowcell.Batch, MappedReads = Mapped.Reads, IntragenicRate = Intragenic.Rate, 
                ExonicRate = Exonic.Rate, IntronicRate = Intronic.Rate, IntergenicRate = Intergenic.Rate, 
                GenesDetected = Genes.Detected, TranscriptsDetected = Transcripts.Detected, 
                ExpProfEfficiency = Expression.Profiling.Efficiency, rRNARate = rRNA.Rate, 
                TotalReads = Total.Reads, PercentAligned = Percent.Aligned)

@@

                          "ACC_RNA_isolation: Exclude?", "ACC_RNA_isolation: Sample RNA ID",
                          "ACC_RNA_isolation: RIN", "ACC_RNA_report: Exclude?",

@@

METADATA_ACC = METADATA %>%
  dplyr::select(Individual_ID, Institution, Gender, Age_of_Death, PMI__in_hours_, Dx, 
                DNA_genotyping_report__Genotyping_Sample_ID, ACC_RNA_isolation__Exclude_, 
                ACC_RNA_isolation__Sample_RNA_ID, ACC_RNA_isolation__RIN, ACC_RNA_report__Exclude_) %>%
  dplyr::rename(PMI = PMI__in_hours_, 
                IsolationExclude = ACC_RNA_isolation__Exclude_, 
                SampleID = ACC_RNA_isolation__Sample_RNA_ID, 
                RIN = ACC_RNA_isolation__RIN, 
                ReportExclude = ACC_RNA_report__Exclude_,
                GenotypingSampleID = DNA_genotyping_report__Genotyping_Sample_ID) %>%
  dplyr::filter(SampleID %in% colnames(COUNT)) %>%
  dplyr::mutate(Tissue = 'ACC')

@@

METADATA = rbindlist(list(METADATA_ACC, METADATA_DLPFC), use.names = T, fill = T) %>%
  droplevels() %>%
  list(.,METADATA_QC) %>%
  join_all %>%
  dplyr::mutate(LibraryBatch = as.factor(LibraryBatch)) %>%
  left_join(ANCESTRY) %>%
  unique

@@

### Outlier analysis
Detect outlier samples based on their expression (logCPM) pattern
```{r detect.outliers, fig.height=8, fig.width=8, results='asis', cache = FALSE}
# Based on PC
indToRemove = c('MSSM_RNA_ACC_BP_24', 'MSSM_RNA_ACC_375', 'MSSM_RNA_ACC_170', 'MSSM_RNA_ACC_315', 'MSSM_RNA_ACC_250',
                'MSSM_RNA_ACC_21', 'MSSM_RNA_ACC_356', 'MSSM_RNA_ACC_318', 'MSSM_RNA_ACC_222', 'MSSM_RNA_ACC_220',
                'MSSM_RNA_ACC_379', 'MSSM_RNA_PFC_133', 'MSSM_RNA_PFC_139', 'MSSM_RNA_PFC_163', 'MSSM_RNA_PFC_260',
                'MSSM_RNA_PFC_299', 'MSSM_RNA_PFC_361', 'MSSM_RNA_PFC_89', 'MSSM_RNA_PFC_319', 'MSSM_RNA_PFC_348',
                'MSSM_RNA_PFC_354', 'MSSM_RNA_PFC_60', 'PENN_RNA_PFC_71', 'PENN_RNA_PFC_56',  "MSSM_RNA_ACC_310",
                "MSSM_RNA_ACC_51",  "MSSM_RNA_ACC_254", "MSSM_RNA_ACC_28", "MSSM_RNA_ACC_293", "MSSM_RNA_ACC_233",
                "MSSM_RNA_ACC_167", "MSSM_RNA_ACC_353", "MSSM_RNA_ACC_355", "MSSM_RNA_ACC_211", "PENN_RNA_ACC_122",
                "PENN_RNA_ACC_049", "PENN_RNA_ACC_052", "PENN_RNA_ACC_068", "PENN_RNA_ACC_070", "PITT_RNA_ACC_10020",
                "PITT_RNA_ACC_1240", "PITT_RNA_ACC_1284", "PITT_RNA_ACC_700", "PITT_RNA_ACC_857", "PITT_RNA_ACC_BP_1589",
                "MSSM_RNA_PFC_230", "MSSM_RNA_PFC_304", "MSSM_RNA_PFC_309", "MSSM_RNA_PFC_329", "MSSM_RNA_PFC_82",
                "PENN_RNA_PFC_5", "PENN_RNA_PFC_58", "PENN_RNA_PFC_99", "PITT_RNA_PFC_1284")

@@

# Estimate voom weights
VOOM.GENE_EXPRESSION = voom(NEW.COUNTS, design = DESIGN)
VOOM.GENE_EXPRESSION$E = CQN.GENE_EXPRESSION$E
correlation <- parallelDuplicateCorrelation(VOOM.GENE_EXPRESSION, block = COVARIATES$Individual_ID)

# Re-calculate voom weights with correlation of random effects
VOOM.GENE_EXPRESSION = voom(NEW.COUNTS,   
                            design = DESIGN, plot=F,
                            block = COVARIATES$Individual_ID, 
                            correlation = correlation$cor)
VOOM.GENE_EXPRESSION$E = CQN.GENE_EXPRESSION$E

@@

### Differential expression analysis (SCZ - Control)
Genes that are differentially expressed at an FDR <= 0.05 are
```{r diffExp, fig.height=15, fig.width=20}
# Fit contrast
contrast = makeContrasts(contrasts=c("Dx.TissueSCZ_ACC-Dx.TissueControl_ACC",
                                     "Dx.TissueSCZ_ACC-Dx.TissueOther_ACC",
                                     "Dx.TissueOther_ACC-Dx.TissueControl_ACC",

                                     "Dx.TissueSCZ_DLPFC-Dx.TissueControl_DLPFC",
                                     "Dx.TissueSCZ_DLPFC-Dx.TissueOther_DLPFC",
                                     "Dx.TissueOther_DLPFC-Dx.TissueControl_DLPFC",

                                     "Dx.TissueSCZ_DLPFC-Dx.TissueSCZ_ACC",
                                     "Dx.TissueControl_DLPFC-Dx.TissueControl_ACC",
                                     "Dx.TissueOther_DLPFC-Dx.TissueOther_ACC",

                                     "Dx.TissueSCZ_DLPFC+Dx.TissueSCZ_ACC-Dx.TissueControl_DLPFC-Dx.TissueControl_ACC",         
                                     "Dx.TissueSCZ_DLPFC+Dx.TissueControl_DLPFC-Dx.TissueSCZ_ACC-Dx.TissueControl_ACC",
                                     "Dx.TissueSCZ_DLPFC-Dx.TissueControl_DLPFC-Dx.TissueSCZ_ACC+Dx.TissueControl_ACC"),
                         levels = colnames(ADJUSTED.FIT$coefficients))
FIT.CONTR = contrasts.fit(ADJUSTED.FIT, contrasts=contrast)
FIT.CONTR = eBayes(FIT.CONTR)

# Get differnetial expression
DE = lapply(1:dim(contrast)[2], function(i, FIT){
  topTable(FIT, coef=i, number = 50000, confint = TRUE) %>%
    rownameToFirstColumn('ensembl_gene_id') 
}, FIT.CONTR) 
names(DE) = gsub('Dx.Tissue', '', colnames(contrast))
names(DE)[10:12] = c('SCZ_ALL-Control_ALL', 'ALL_DLPFC-ALL_ACC', 'ALL_ALL-ALL_ALL')

@@

p = DE %>%
  dplyr::filter(ref.Dx != 'ALL', to.Dx != 'ALL') %>%
  ggplot(aes(y = -log10(adj.P.Val), x = logFC, color = Direction)) + geom_point() + xlim(c(-1,1))
p = p + scale_color_manual(values = c('green','grey','red')) + theme(legend.position = 'top')
p = p + facet_grid(ref.Dx+ref.Tissue~.+to.Dx+to.Tissue, scales = 'fixed')
p = p + ggtitle('Differential Expression between Schizophrenia and Control')
p

p = DE %>%
  dplyr::filter(ref.Dx == 'ALL', to.Dx == 'ALL') %>%
  ggplot(aes(y = -log10(adj.P.Val), x = logFC, color = Direction)) + geom_point() + xlim(c(-1,1))
p = p + scale_color_manual(values = c('green','grey','red')) + theme(legend.position = 'top')
p = p + facet_grid(ref.Tissue+.~to.Tissue, scales = 'fixed')
p = p + ggtitle('Differential Expression between DLPFC and ACC')
p
```

@@

# Estimate voom weights
VOOM.GENE_EXPRESSION1 = voom(NEW.COUNTS, design = DESIGN1)
VOOM.GENE_EXPRESSION1$E = CQN.GENE_EXPRESSION$E
correlation <- parallelDuplicateCorrelation(VOOM.GENE_EXPRESSION1, block = COVARIATES$Individual_ID)

# Re-calculate voom weights with correlation of random effects
VOOM.GENE_EXPRESSION1 = voom(NEW.COUNTS,   
                             design = DESIGN1, plot=F,
                             block = COVARIATES$Individual_ID, 
                             correlation = correlation$cor)
VOOM.GENE_EXPRESSION1$E = CQN.GENE_EXPRESSION$E

@@

# Fit contrast
contrast = makeContrasts(contrasts=c("Dx.Tissue.GenderSCZ_ACC_Male-Dx.Tissue.GenderControl_ACC_Male",
                                     "Dx.Tissue.GenderSCZ_ACC_Female-Dx.Tissue.GenderControl_ACC_Female",

                                     "Dx.Tissue.GenderSCZ_DLPFC_Male-Dx.Tissue.GenderControl_DLPFC_Male",
                                     "Dx.Tissue.GenderSCZ_DLPFC_Female-Dx.Tissue.GenderControl_DLPFC_Female"),
                         levels = colnames(ADJUSTED.FIT1$coefficients))
FIT.CONTR1 = contrasts.fit(ADJUSTED.FIT1, contrasts=contrast)
FIT.CONTR1 = eBayes(FIT.CONTR1)

@@

### Differential expression analysis (ACC - DLPFC, SCZ - Control, Female - Male)
Genes that are differentially expressed at an FDR <= 0.05 are
```{r diffExp2, fig.height=16, fig.width=8}
COVARIATES = tidyr::separate(COVARIATES, Dx.Tissue, c('Dx', 'Tissue')) %>%
  dplyr::mutate(Dx = factor(Dx, levels = c('SCZ', 'Control', 'Other')),
                Tissue = factor(Tissue))
postAdjustCovars = c('Tissue', 'Dx', 'Institution', 'Gender', 'RIN', 'IntronicRate')

# Post adjusted design matrix
DESIGN2 = getDesignMatrix(COVARIATES[,postAdjustCovars,drop=F],Intercept = F)$design
DESIGN2 = DESIGN2[,linColumnFinder(DESIGN2)$indepCols]

# Estimate voom weights
VOOM.GENE_EXPRESSION2 = voom(NEW.COUNTS, design = DESIGN2)
VOOM.GENE_EXPRESSION2$E = CQN.GENE_EXPRESSION$E
correlation <- parallelDuplicateCorrelation(VOOM.GENE_EXPRESSION2, block = COVARIATES$Individual_ID)

# Re-calculate voom weights with correlation of random effects
VOOM.GENE_EXPRESSION2 = voom(NEW.COUNTS,   
                             design = DESIGN2, plot=F,
                             block = COVARIATES$Individual_ID, 
                             correlation = correlation$cor)
VOOM.GENE_EXPRESSION2$E = CQN.GENE_EXPRESSION$E

# Fit linear model using new weights and new design
ADJUSTED.FIT2 = lmFit(VOOM.GENE_EXPRESSION2)

# Fit contrast
contrast = makeContrasts(contrasts=c("TissueACC-TissueDLPFC",
                                     "DxSCZ-DxControl",
                                     "GenderFemale"),
                         levels = colnames(ADJUSTED.FIT2$coefficients))
FIT.CONTR2 = contrasts.fit(ADJUSTED.FIT2, contrasts=contrast)
FIT.CONTR2 = eBayes(FIT.CONTR2)

# Get differnetial expression
DE2 = lapply(1:dim(contrast)[2], function(i, FIT){
  topTable(FIT, coef=i, number = 50000, confint = TRUE) %>%
    rownameToFirstColumn('ensembl_gene_id') 
}, FIT.CONTR2) 
names(DE2) = c('ACC-DLPFC', 'SCZ-CONTROL', 'FEMALE-MALE')

DE2 = DE2 %>% 
  rbindlist(idcol = 'Comparison') %>%
  dplyr::mutate(Direction = sign(logFC),
                Direction = factor(Direction, c(-1,1), c('-1' = 'DOWN', '1' = 'UP')),
                Direction = as.character(Direction)) %>%
  left_join(Ensemble2HGNC) 
DE2$Direction[DE2$adj.P.Val > 0.05 | abs(DE2$logFC) < log2(1.2)] = 'NONE'

tmp = DE2 %>%
  dplyr::filter(adj.P.Val <= 0.05) %>%
  dplyr::select(Comparison, ensembl_gene_id, logFC) %>%
  group_by(Comparison) %>%
  dplyr::summarise(FDR_0_05_DOWN = length(unique(ensembl_gene_id[logFC < 0])),
                   FDR_0_05_UP = length(unique(ensembl_gene_id[logFC > 0])))

tmp1 = DE2 %>%
  dplyr::filter(adj.P.Val <= 0.05, abs(logFC) >= log2(1.2)) %>%
  dplyr::select(Comparison, ensembl_gene_id, logFC) %>%
  group_by(Comparison) %>%
  dplyr::summarise(FDR_0_05_FC_1.2_DOWN = length(unique(ensembl_gene_id[logFC < 0])),
                   FDR_0_05_FC_1.2_UP = length(unique(ensembl_gene_id[logFC > 0])))

p = DE2 %>%
  ggplot(aes(y = -log10(adj.P.Val), x = logFC, color = Direction)) + geom_point() + xlim(c(-1,1))
p = p + scale_color_manual(values = c('green','grey','red')) + theme(legend.position = 'top')
p = p + facet_grid(Comparison~., scales = 'free_y')
p = p + ggtitle('Differential Expression')
p
```

### Gene set enrichment analysis
Using Fisher's exact test for gene set enrichment analysis
```{r enrich.de, fig.height=16, fig.width=16}
all.diff.exp = list(DE %>%
                      tidyr::unite(Comparison1, to.Dx, to.Tissue) %>%
                      tidyr::unite(Comparison2, ref.Dx, ref.Tissue) %>%
                      tidyr::unite(Comparison, Comparison1, Comparison2, sep = '-'), 
                    DE1%>%
                      tidyr::unite(Comparison1, to.Dx, to.Tissue, to.Gender) %>%
                      tidyr::unite(Comparison2, ref.Dx, ref.Tissue, ref.Gender) %>%
                      tidyr::unite(Comparison, Comparison1, Comparison2, sep = '-'), 
                    DE2) %>%
  data.table::rbindlist(fill = TRUE, use.names = TRUE)
diffexp.genes.up = all.diff.exp %>%
  dlply(.(Comparison), .fun = function(x){
    tmp = x$hgnc_symbol[x$adj.P.Val <= 0.05 & x$logFC > 0] %>% unique
    tmp = tmp[!is.na(tmp)]
  })
names(diffexp.genes.up) = paste(names(diffexp.genes.up), 'UP', sep = '.')

@@

  tissueTypeAbrv	= 'DLPFC, ACC', 
  study = 'CMC',

@@

# Code
CODE <- Folder(name = "ACC and DLPFC - STAR - FEATURE COUNTS - ENSEMBL v75 - CQN", parentId = parentId)
annotations(CODE) = all.annotations
CODE <- synStore(CODE)

# Store covariates
COVARIATES = rownameToFirstColumn(COVARIATES, 'SampleID')
write.table(COVARIATES, file = 'CMC_ACC_DLPFC_Covariates.tsv', sep = '\t', row.names=F, quote=F)
COV_OBJ = File('CMC_ACC_DLPFC_Covariates.tsv', name = 'Covariates', parentId = CODE$properties$id)
annotations(COV_OBJ) = annotations(CODE)
annotations(COV_OBJ)$dataSubType = 'covariates'
COV_OBJ = synStore(COV_OBJ, used = ALL_USED_IDs, activityName = activityName, 
                      executed = thisFile, activityDescription = activityDescription)

# Store filtered counts
PROCESSED_COUNTS$filteredExprMatrix$counts %>%
  rownameToFirstColumn('ensembl_gene_id') %>%
  write.table(file = 'CMC_ACC_DLPFC_Counts.tsv', sep = '\t', row.names=F, quote=F)
COUNT_OBJ = File('CMC_ACC_DLPFC_Counts.tsv', name = 'Counts (filtered raw)', parentId = CODE$properties$id)
annotations(COUNT_OBJ) = annotations(CODE)
annotations(COUNT_OBJ)$dataSubType = 'filteredCounts'
COUNT_OBJ = synStore(COUNT_OBJ, activity = synGetActivity(COV_OBJ$properties$id))

# Store logCPM
CQN.GENE_EXPRESSION$y[is.na(CQN.GENE_EXPRESSION$E)] = NA
CQN.GENE_EXPRESSION$y %>%
  rownameToFirstColumn('ensembl_gene_id') %>%
  write.table(file = 'CMC_ACC_DLPFC_logCPM.tsv', sep = '\t', row.names=F, quote=F)
LCOUNT_OBJ = File('CMC_ACC_DLPFC_logCPM.tsv', name = 'Counts (filtered logCPM)', parentId = CODE$properties$id)
annotations(LCOUNT_OBJ) = annotations(CODE)
annotations(LCOUNT_OBJ)$dataSubType = 'filteredLCPM'
LCOUNT_OBJ = synStore(LCOUNT_OBJ, activity = synGetActivity(COV_OBJ$properties$id))

# Store cqn offsets
CQN.GENE_EXPRESSION$offset[is.na(CQN.GENE_EXPRESSION$E)] = NA
CQN.GENE_EXPRESSION$offset %>%
  rownameToFirstColumn('ensembl_gene_id') %>%
  write.table(file = 'CMC_ACC_DLPFC_offset.tsv', sep = '\t', row.names=F, quote=F)
OFFSET_OBJ = File('CMC_ACC_DLPFC_offset.tsv', name = 'Gene length and GC content offset', parentId = CODE$properties$id)
annotations(OFFSET_OBJ) = annotations(CODE)
annotations(OFFSET_OBJ)$dataSubType = 'offset'
OFFSET_OBJ = synStore(OFFSET_OBJ, activity = synGetActivity(COV_OBJ$properties$id))

# Store design matrix
DESIGN %>%
  rownameToFirstColumn('SampleID') %>%
  write.table(file = 'CMC_ACC_DLPFC_Design.tsv', sep = '\t', row.names=F, quote=F)
DM_OBJ = File('CMC_ACC_DLPFC_Design.tsv', name = 'Design Matrix', parentId = CODE$properties$id)
annotations(DM_OBJ) = annotations(CODE)
annotations(DM_OBJ)$dataSubType = 'designMatrix'
DM_OBJ = synStore(DM_OBJ, activity = synGetActivity(COV_OBJ$properties$id))

>>>>

---
title: "Covariate analysis of CMC RNASeq data reprocessed with STAR-feature counts workflow with Ensembl v75 (DLPFC with CQN normlaisation)"
author: "Thanneer Perumal"
date: "`r date()`"
output: html_document
---
```{r knit2synapse, eval=FALSE}
library(synapseClient)
library(knit2synapse)

@@

knit2synapse::knitToFolderEntity(file = "./CMC_DLPFC_STAR_ENSEMBL_v75_CQN.Rmd",
                                 parentId = "syn7501880",
                                 entityName = 'DLPFC - STAR - FEATURE COUNTS - ENSEMBL v75 - CQN')
```

@@

```{r synapse.parameters, include=FALSE, cache=TRUE}
parentId = "syn7501880";
activityName = 'Covariate adjustments';
activityDescription = 'Covariate analysis of STAR aligned feature counts with Ensembl v75 with CQN normalisation (DLPFC)';

thisFileName <- 'CMC_DLPFC_STAR_ENSEMBL_v75_CQN.Rmd'

# Github link
thisRepo <- getRepo(repository = "th1vairam/Brain_Reg_Net", ref="branch", refName='CMC')
thisFile <- getPermlink(repository = thisRepo, repositoryPath=paste0('code/Rmd/',thisFileName))
```
### Data download
Obtain count matrix and metadata from synapse.
```{r download.data, cache=TRUE}
# Download reprocessed counts (DLPFC)
COUNT_DLPFC_ID = 'syn8413221';
ALL_USED_IDs = COUNT_DLPFC_ID
COUNT = downloadFile(COUNT_DLPFC_ID) %>%
  as.data.frame()

GENE.LEN = dplyr::select(COUNT, Geneid, Length) %>%
  dplyr::rename(ensembl_gene_id = Geneid) %>%
  unique()
rownames(GENE.LEN) = GENE.LEN$ensembl_gene_id

@@

# Get RNA qc metrics from synapse
METADATA_QC_DLPFC_ID = 'syn2279446'
ALL_USED_IDs[length(ALL_USED_IDs)+1] = METADATA_QC_DLPFC_ID
METADATA_QC_DLPFC = read.table(synGet(METADATA_QC_DLPFC_ID)@filePath, fill = T, header = T, sep = ',') %>%
  dplyr::filter(is.na(Exclude.)) %>%
  dplyr::select(Sample.RNA.ID, Ribozero.Batch, Library.Batch, Flowcell.Batch, Mapped.Reads, 
                Intragenic.Rate, Exonic.Rate, Intronic.Rate, Intergenic.Rate, Genes.Detected, 
                Transcripts.Detected, Expression.Profiling.Efficiency, rRNA.Rate, Total.Reads, Percent.Aligned) %>%
  dplyr::rename(SampleID = Sample.RNA.ID, RibozeroBatch = Ribozero.Batch, LibraryBatch = Library.Batch, 
                FlowcellBatch = Flowcell.Batch, MappedReads = Mapped.Reads, IntragenicRate = Intragenic.Rate, 
                ExonicRate = Exonic.Rate, IntronicRate = Intronic.Rate, IntergenicRate = Intergenic.Rate, 
                GenesDetected = Genes.Detected, TranscriptsDetected = Transcripts.Detected, 
                ExpProfEfficiency = Expression.Profiling.Efficiency, rRNARate = rRNA.Rate, 
                TotalReads = Total.Reads, PercentAligned = Percent.Aligned)

@@

@@

@@

METADATA = METADATA_DLPFC %>%
  droplevels() %>%
  list(.,METADATA_QC_DLPFC) %>%
  join_all %>%
  dplyr::mutate(LibraryBatch = as.factor(LibraryBatch)) %>%
  left_join(ANCESTRY) %>%
  unique

@@

### Outlier analysis
Detect outlier samples based on their expression (logCPM) pattern
```{r detect.outliers, fig.height=8, fig.width=8, results='asis', cache = FALSE}
# Based on PC
indToRemove = c('MSSM_RNA_PFC_133', 'MSSM_RNA_PFC_139', 'MSSM_RNA_PFC_163', 'MSSM_RNA_PFC_260',
                'MSSM_RNA_PFC_299', 'MSSM_RNA_PFC_361', 'MSSM_RNA_PFC_89', 'MSSM_RNA_PFC_319', 'MSSM_RNA_PFC_348',
                'MSSM_RNA_PFC_354', 'MSSM_RNA_PFC_60', 'PENN_RNA_PFC_71', 'PENN_RNA_PFC_56',  "MSSM_RNA_PFC_230",
                "MSSM_RNA_PFC_304", "MSSM_RNA_PFC_309", "MSSM_RNA_PFC_329", "MSSM_RNA_PFC_82",
                "PENN_RNA_PFC_5", "PENN_RNA_PFC_58", "PENN_RNA_PFC_99", "PITT_RNA_PFC_1284")

@@

# Estimate voom weights
VOOM.GENE_EXPRESSION = voom(NEW.COUNTS, design = DESIGN)
VOOM.GENE_EXPRESSION$E = CQN.GENE_EXPRESSION$E

@@

### Differential expression analysis (SCZ - Control)
Genes that are differentially expressed at an FDR <= 0.05 are
```{r diffExp, fig.height=15, fig.width=20}
# Fit contrast
contrast = makeContrasts(contrasts=c("Dx.TissueSCZ_DLPFC-Dx.TissueControl_DLPFC",
                                     "Dx.TissueSCZ_DLPFC-Dx.TissueOther_DLPFC",
                                     "Dx.TissueOther_DLPFC-Dx.TissueControl_DLPFC"),
                         levels = colnames(ADJUSTED.FIT$coefficients))
FIT.CONTR = contrasts.fit(ADJUSTED.FIT, contrasts=contrast)
FIT.CONTR = eBayes(FIT.CONTR)

# Get differnetial expression
DE = lapply(1:dim(contrast)[2], function(i, FIT){
  topTable(FIT, coef=i, number = 50000, confint = TRUE) %>%
    rownameToFirstColumn('ensembl_gene_id') 
}, FIT.CONTR) 
names(DE) = gsub('Dx.Tissue', '', colnames(contrast))

@@

p = DE %>%
  dplyr::filter(ref.Dx != 'ALL', to.Dx != 'ALL') %>%
  ggplot(aes(y = -log10(adj.P.Val), x = logFC, color = Direction)) + geom_point() + xlim(c(-1,1))
p = p + scale_color_manual(values = c('green','grey','red')) + theme(legend.position = 'top')
p = p + facet_grid(ref.Dx+ref.Tissue~.+to.Dx+to.Tissue, scales = 'fixed')
p = p + ggtitle('Differential Expression between Schizophrenia and Control')
p
```

@@

# Estimate voom weights
VOOM.GENE_EXPRESSION1 = voom(NEW.COUNTS, design = DESIGN1)
VOOM.GENE_EXPRESSION1$E = CQN.GENE_EXPRESSION$E

@@

# Fit contrast
contrast = makeContrasts(contrasts=c("Dx.Tissue.GenderSCZ_DLPFC_Male-Dx.Tissue.GenderControl_DLPFC_Male",
                                     "Dx.Tissue.GenderSCZ_DLPFC_Female-Dx.Tissue.GenderControl_DLPFC_Female"),
                         levels = colnames(ADJUSTED.FIT1$coefficients))
FIT.CONTR1 = contrasts.fit(ADJUSTED.FIT1, contrasts=contrast)
FIT.CONTR1 = eBayes(FIT.CONTR1)

@@

### Gene set enrichment analysis
Using Fisher's exact test for gene set enrichment analysis
```{r enrich.de, fig.height=16, fig.width=16}
all.diff.exp = list(DE %>%
                      tidyr::unite(Comparison1, to.Dx, to.Tissue) %>%
                      tidyr::unite(Comparison2, ref.Dx, ref.Tissue) %>%
                      tidyr::unite(Comparison, Comparison1, Comparison2, sep = '-'), 
                    DE1%>%
                      tidyr::unite(Comparison1, to.Dx, to.Tissue, to.Gender) %>%
                      tidyr::unite(Comparison2, ref.Dx, ref.Tissue, ref.Gender) %>%
                      tidyr::unite(Comparison, Comparison1, Comparison2, sep = '-')) %>%
  data.table::rbindlist(fill = TRUE, use.names = TRUE)
diffexp.genes.up = all.diff.exp %>%
  dlply(.(Comparison), .fun = function(x){
    tmp = x$hgnc_symbol[x$adj.P.Val <= 0.05 & x$logFC > 0] %>% unique
    tmp = tmp[!is.na(tmp)]
  })
names(diffexp.genes.up) = paste(names(diffexp.genes.up), 'UP', sep = '.')

@@

  tissueTypeAbrv	= 'DLPFC', 
  study = 'CMC',

@@

# Code
CODE <- Folder(name = "DLPFC - STAR - FEATURE COUNTS - ENSEMBL v75 - CQN", parentId = parentId)
annotations(CODE) = all.annotations
CODE <- synStore(CODE)

# Store covariates
COVARIATES = rownameToFirstColumn(COVARIATES, 'SampleID')
write.table(COVARIATES, file = 'CMC_DLPFC_Covariates.tsv', sep = '\t', row.names=F, quote=F)
COV_OBJ = File('CMC_DLPFC_Covariates.tsv', name = 'Covariates', parentId = CODE$properties$id)
annotations(COV_OBJ) = annotations(CODE)
annotations(COV_OBJ)$dataSubType = 'covariates'
COV_OBJ = synStore(COV_OBJ, used = ALL_USED_IDs, activityName = activityName, 
                      executed = thisFile, activityDescription = activityDescription)

# Store filtered counts
PROCESSED_COUNTS$filteredExprMatrix$counts %>%
  rownameToFirstColumn('ensembl_gene_id') %>%
  write.table(file = 'CMC_DLPFC_Counts.tsv', sep = '\t', row.names=F, quote=F)
COUNT_OBJ = File('CMC_DLPFC_Counts.tsv', name = 'Counts (filtered raw)', parentId = CODE$properties$id)
annotations(COUNT_OBJ) = annotations(CODE)
annotations(COUNT_OBJ)$dataSubType = 'filteredCounts'
COUNT_OBJ = synStore(COUNT_OBJ, activity = synGetActivity(COV_OBJ$properties$id))

# Store logCPM
CQN.GENE_EXPRESSION$y[is.na(CQN.GENE_EXPRESSION$E)] = NA
CQN.GENE_EXPRESSION$y %>%
  rownameToFirstColumn('ensembl_gene_id') %>%
  write.table(file = 'CMC_DLPFC_logCPM.tsv', sep = '\t', row.names=F, quote=F)
LCOUNT_OBJ = File('CMC_DLPFC_logCPM.tsv', name = 'Counts (filtered logCPM)', parentId = CODE$properties$id)
annotations(LCOUNT_OBJ) = annotations(CODE)
annotations(LCOUNT_OBJ)$dataSubType = 'filteredLCPM'
LCOUNT_OBJ = synStore(LCOUNT_OBJ, activity = synGetActivity(COV_OBJ$properties$id))

# Store cqn offsets
CQN.GENE_EXPRESSION$offset[is.na(CQN.GENE_EXPRESSION$E)] = NA
CQN.GENE_EXPRESSION$offset %>%
  rownameToFirstColumn('ensembl_gene_id') %>%
  write.table(file = 'CMC_DLPFC_offset.tsv', sep = '\t', row.names=F, quote=F)
OFFSET_OBJ = File('CMC_DLPFC_offset.tsv', name = 'Gene length and GC content offset', parentId = CODE$properties$id)
annotations(OFFSET_OBJ) = annotations(CODE)
annotations(OFFSET_OBJ)$dataSubType = 'offset'
OFFSET_OBJ = synStore(OFFSET_OBJ, activity = synGetActivity(COV_OBJ$properties$id))

# Store design matrix
DESIGN %>%
  rownameToFirstColumn('SampleID') %>%
  write.table(file = 'CMC_DLPFC_Design.tsv', sep = '\t', row.names=F, quote=F)
DM_OBJ = File('CMC_DLPFC_Design.tsv', name = 'Design Matrix', parentId = CODE$properties$id)
annotations(DM_OBJ) = annotations(CODE)
annotations(DM_OBJ)$dataSubType = 'designMatrix'
DM_OBJ = synStore(DM_OBJ, activity = synGetActivity(COV_OBJ$properties$id))

# Store residual gene expression for network analysis
RESIDUAL.GENE_EXPRESSION %>%
  rownameToFirstColumn('ensembl_gene_id') %>%
  write.table(file = 'CMC_DLPFC_netResidualExpression.tsv', sep = '\t', row.names=F, quote=F)
nEXP_OBJ = File('CMC_DLPFC_netResidualExpression.tsv', 
                name = 'Normalised, covariates removed residual expression (for network analysis)', 
                parentId = CODE$properties$id)
annotations(nEXP_OBJ) = annotations(CODE)
annotations(nEXP_OBJ)$dataSubType = 'residualGeneExpForNetAnlz'
nEXP_OBJ = synStore(nEXP_OBJ, activity = synGetActivity(COV_OBJ$properties$id))

# Store differential expression results
write.table(all.diff.exp, file = 'CMC_DLPFC_DiffExpression.tsv', sep = '\t', row.names=F, quote=F)
DEXP_OBJ = File('CMC_DLPFC_DiffExpression.tsv', 
                name = 'Differential Expression Results (Dx.Tissue)', 
                parentId = CODE$properties$id)
annotations(DEXP_OBJ) = annotations(CODE)
annotations(DEXP_OBJ)$dataSubType = 'diffExp'
DEXP_OBJ = synStore(DEXP_OBJ, activity = synGetActivity(COV_OBJ$properties$id))
